Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-25 trunk/paper/drinking.tex

\subsection{Processes, their participants and qualities}
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{onto.pdf}
  \caption{Overview over ...\label{fig:onto}}
\end{figure}

@@

\section{Results}
\subsection{Attributes of processes}
An attribute that can be directly linked to a process is its {\em
  duration}. A duration is an attribute that characterizes the
temporal extent of a process and is similar to {\em Length}, {\em
  Area} and {\em Volume} for one-, two- and three-dimensional
entities. Furthermore, processes may be attributed a {\em temporal
  location} with respect to a particular coordinate system.

@@

\section{Discussion}
\subsection{Implementation as pre-composed ontology}
Pre-composed or software support for pattern?

\subsection{Use of CPO}

The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO BP 'cellular process' as the entity; a further 11\%
reference GO CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration}, {\em decreased
  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
and stands to benefit from the automated synchronization between CPO
and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
body of highly specific, well-supported data to be integrated with
data from other species.

\subsection{Future research: acts and intentional states}
activities/actions, intentional (goal-oriented) acts

feeling thirsty (internal state) as driving force behind drinking
behavior

non-measurable, inference only

>>>>

\subsection{Processes and their participants}
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{onto.pdf}
  \caption{Overview over ...\label{fig:onto}}
\end{figure}

@@

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes:
the first are process attributes that arise directly from processes
and include {\em duration} and {\em temporal location}; the second are
attributes that arise from processes and their parts and include {\em
  frequency}; and the third are attributes that arise from processes
and qualities of their participants, and include {\em flow rates}.

Attribute that can be directly linked to a process are the result of
the temporal extension of a process and include process {\em duration}
and {\em temporal location}. A duration is an attribute that
characterizes the temporal extent of a process and is similar to {\em
  Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional entities. A {\em Temporal location} positions the
time interval during which the process occurs with respect to a
reference coordinate system.

@@

\section{Discussion}
\subsection{Use of CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO BP 'cellular process' as the entity; a further 11\%
reference GO CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration}, {\em decreased
  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
and stands to benefit from the automated synchronization between CPO
and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
body of highly specific, well-supported data to be integrated with
data from other species.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
A domain that would greatly benefit from the development of a CPO is
Systems Microscopy, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
(Lock \& Stromblad, Exp. Cell Res. 2010).

\subsection{Pre- vs. post-composed ontology}
Pre-composed or software support for pattern?
